---
title: "Part 1"
output: html_document
date: "2026-08-12"
---

# setting up packages
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
install.packages("dplyr")
install.packages("tidyverse")
install.packages("janitor")
install.packages("lubridate")
library(dplyr)
library(tidyverse)
library(janitor)
library(lubridate)
```

# reading in juvenile arrest data
```{r}
juvenile_arrests <- read_csv("data/Juvenile_Arrests.csv")
```

# basic cleaning on column names, then collapses white space and turns characters lowercase
```{r}
juvenile_arrests <- juvenile_arrests %>%
  clean_names() %>%
  mutate(across(where(is.character), ~ str_to_lower(str_squish(.x))))
```

# fixing arrest date column into real dates
```{r}
juvenile_arrests <- juvenile_arrests %>%
  mutate(arrest_date = as_date(ymd_hms(arrest_date)))
```

# visually checking data coverage, are there values for all months for each year?
```{r}
monthly_counts <- juvenile_arrests %>%
  mutate(year_month = floor_date(arrest_date, "month")) %>%
  count(year_month, name = "total_arrests") %>%
  arrange(year_month)
```

## There's good data coverage for all months from 2011 to 2024. I think it would be interesting to see if there is any relationship between juvenile arrests and the month of the year, to then see if the school year impacts arrests. 

# extracting the month from the dates and getting counts
```{r}
monthly_totals <- juvenile_arrests %>%
  mutate(month = month(arrest_date, label = TRUE, abbr = TRUE)) %>%
  group_by(month) %>%
  summarize(total_arrests = n(), .groups = "drop") %>%
  arrange(desc(total_arrests))

print(monthly_totals)
```

# chart
```{r}
ggplot(monthly_totals, aes(x = month, y = total_arrests)) +
  geom_col(fill = "blue") +
  geom_text(aes(label = scales::comma(total_arrests)), vjust = -0.5, size = 3.5) +
  labs(
    title = "Total Juvenile Arrests by Month (All Years Combined)",
    x = "Month",
    y = "Total Arrests"
  ) +
  theme_minimal() +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.1)))
```


# I don't see a strong relationship between season and number of juvenile arrests, except for a November/December decrease. I'll move on to looking at counts per year to see if there have been overall changes in the number of arrests over time.

# getting yearly totals
```{r}
yearly_totals <- juvenile_arrests %>%
  mutate(year = year(arrest_date)) %>%
  group_by(year) %>%
  summarize(total_arrests = n(), .groups = "drop") %>%
  arrange(year)

print(yearly_totals)

#simple chart
yearly_totals %>%
  ggplot(aes(year, total_arrests)) +
  geom_col(fill = "steelblue") +
  labs(title = "Total Arrests by Year", x = "Year", y = "Number of Arrests")
```

# adding percent change between all years
```{r}
yearly_totals <- yearly_totals %>%
  mutate(
    pct_change = (total_arrests - lag(total_arrests)) / lag(total_arrests) * 100)

(last(yearly_totals$total_arrests) - first(yearly_totals$total_arrests)) / first(yearly_totals$total_arrests) * 100
```


## Juvenile arrests have decreased by 50% between 2011 and 2024. The steepest decrease was between 2019 and 2020. While arrest totals have fluctuated since 2020, they have not returned to pre-pandemic levels.

# final viz, line chart with yearly totals
```{r}
line_chart <- ggplot(yearly_totals, aes(x = year, y = total_arrests, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  geom_text(aes(label = scales::comma(total_arrests)), vjust = -0.8, size = 3.5) +
  labs(
    title = "Total Juvenile Arrests per Year in Washington, D.C.",
    x = "Year",
    y = "Total Arrests"
  ) +
  theme_minimal() +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.15)))

ggsave("yearly_arrests.png", plot = line_chart, width = 8, height = 5, dpi = 300, bg = "white")
```

