---
title: "Part 1"
output: html_document
date: "2026-08-12"
---

# setting up packages
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
install.packages("dplyr")
install.packages("tidyverse")
install.packages("janitor")
install.packages("lubridate")
library(dplyr)
library(tidyverse)
library(janitor)
library(lubridate)
```

# reading in juvenile arrest data
```{r}
juvenile_arrests <- read_csv("data/Juvenile_Arrests.csv")
```

# basic cleaning on column names, then collapses white space and turns characters lowercase
```{r}
juvenile_arrests <- juvenile_arrests %>%
  clean_names() %>%
  mutate(across(where(is.character), ~ str_to_lower(str_squish(.x))))
```

# fixing arrest date column into real dates
```{r}
juvenile_arrests <- juvenile_arrests %>%
  mutate(arrest_date = as_date(ymd_hms(arrest_date)))
```

# visually checking data coverage, are there values for all months for each year?
```{r}
monthly_counts <- juvenile_arrests %>%
  mutate(year_month = floor_date(arrest_date, "month")) %>%
  count(year_month, name = "total_arrests") %>%
  arrange(year_month)
```

## There's good data coverage for all months from 2011 to 2024. I think it would be interesting to see if there is any relationship between juvenile arrests and the month of the year, to then see if the school year impacts arrests. 

# extracting the month from the dates and getting counts
```{r}
monthly_totals <- juvenile_arrests %>%
  mutate(month = month(arrest_date, label = TRUE, abbr = TRUE)) %>%
  group_by(month) %>%
  summarize(total_arrests = n(), .groups = "drop") %>%
  arrange(desc(total_arrests))

print(monthly_totals)
```

# chart
```{r}
ggplot(monthly_totals, aes(x = month, y = total_arrests)) +
  geom_col(fill = "blue") +
  geom_text(aes(label = scales::comma(total_arrests)), vjust = -0.5, size = 3.5) +
  labs(
    title = "Total Juvenile Arrests by Month (All Years Combined)",
    x = "Month",
    y = "Total Arrests"
  ) +
  theme_minimal() +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.1)))
```


# I don't see a strong relationship between season and number of juvenile arrests, except for a November/December decrease. I'll move on to looking at counts per year to see if there have been overall changes in the number of arrests over time.

```{r}

```


